---
title: "ian_project_2"
author: "ian johnson"
date: "March 6, 2016"
output: html_document
---

```{r setup, include=FALSE}
# Initial setup
library(ncdf4)
library(ggplot2)
library(devtools)
library(ggbiplot)
install_github("vqv/ggbiplot")
setwd('~/Desktop/Columbia/EDAV/Project2')
```

```{r, echo=FALSE}
# Load Data
ncname <- "NOAA_Daily_phi_500mb"
ncfname <- paste(ncname, ".nc", sep = "")
dname <- "phi" #For variable name phi
ncin <- nc_open(ncfname)
```

```{r, echo=FALSE}
# Clean data
lon <- ncvar_get(ncin, "X")
lat <- ncvar_get(ncin, "Y", verbose = F)
time_var <- ncvar_get(ncin, "T")
tunits <- ncatt_get(ncin, "T", "units")

# Create 3D array of Phi values (Chris's method)
dmissing_value <- ncatt_get(ncin, dname, "missing_value")
phi.array <- ncvar_get(ncin, dname)
phi.array[phi.array == dmissing_value$value] <- NA

# Set up lat and long lines
x = rep(0:143 * 2.5 - 180, times=15)
y = rep(0:14 * 2.5 + 35, each=144)
# Take phi data from first day
z = phi.array[2161:4320]
df = data.frame(y, x, z)
```





# Phi values overview
Here we show color coded phi values on a spatial grid, with darker colors indicating higher values. We can see that higher lattitudes are associated with higher phi values.

```{r, echo=FALSE}
map.dat <- map_data("world")
ggplot() + geom_polygon(aes(long, lat, group=group), fill="grey65", data=map.dat) +
theme_bw() + theme(axis.text = element_blank(), axis.title=element_blank()) +
geom_point(aes(x=x, y=y, show_guide=TRUE, color=z, size=13), data=df, alpha=.7, na.rm=T) +
scale_color_gradient(low='beige', high='blue')
```




# PCA analysis of lattitude vs longitude vs geopotential height
We next would like some insight into the relationship between geopotential height and lattitude and longitude coordinates. We do PCA analysis to determine which of these three factors accounts for the most variation in the data. We find that longitude seems to be the highest contributor. 

```{r, echo=FALSE}
# PCA analysis of geopotential height data from Jan 1, 1944
colnames(df) = c('Lat', 'Lon', 'Phi')
pca = princomp(df, scores=TRUE, cor=TRUE, center=TRUE, scale=TRUE)
summary(pca)
plot(pca)
biplot(pca, expand=1, xlim=c(-0.05, 0.05), ylim=c(-0.05, 0.05)) 
# todo - use ggbiplot. longitude and phi values are more coorelated than lat and phi...?
pca$loadings
```






```{r, echo=FALSE}
# Load Flood Data (Woojin's method)
df2 = read.csv('GlobalFloodsRecord.csv', fileEncoding="macroman", stringsAsFactors = FALSE)
# Fix dataset
fix.index = c(2,8,22,30)
colnames(df2)[fix.index] = c("Annual.DFO", "Detailed.Locations", "News", "Notes")
num.cols = c(2,12,13,14,15,17,18,19,20,21)
df2[,num.cols] <- sapply(df2[,num.cols], as.numeric)
df2$X <- NULL

# make centroid x 0 -> 360
df2$Centroid.X = df2$Centroid.X + 180
# format dates (just using begin dates for now)
df2$newDate <- as.Date(df2$Began, "%e-%b-%y")
# turn into days since 1948-01-01
Diff <- function(x) as.numeric(x - as.Date('1945-01-01'))
df2 = transform(df2, numDays = Diff(newDate))
```






## Comparing geopotential height values during flooding,
## to average geopotential height values

Next we would like to determine whether higher Geopotential Height values might be coorelated with increased flooding. Here we calculate the means of the phi values during flood events, and compare them against the mean phi value overall. We find that the mean of the geopotential height values during flooding is larger than the mean of geopotential heights overall, but only by a small amount.

```{r, echo=FALSE}
# Take the average of the phi values from days when flooding started
flood_phi_values = c()
for (i in 1:4356) {
  centroid_x = df2$Centroid.X[i]
  centroid_y = df2$Centroid.Y[i]
  lon_index = floor(df2$Centroid.X[i] / 2.5)
  lat_index = floor(df2$Centroid.Y[i] / 2.5)
  if (!is.na(lat_index) && !is.na(lon_index)) {
    if (lat_index > 0 && lat_index < 15) {
      if (!is.na(phi.array[lon_index, lat_index, i])) {
        flood_phi_values[i] = phi.array[lon_index, lat_index, i]
      }
    }
  }
}

# Compare these values to average phi
gh_phi_mean = round(mean(flood_phi_values, na.rm=TRUE), digits=2)
phi_mean = round(mean(df$Phi), digits=2)
gh_min = round(min(df$Phi), digits=2)
gh_max = round(max(df$Phi), digits=2)
```

 Geopotential Height Min Value    Geopotential Height Max Value    Geopotential Height Mean Value    Geopotential Height Mean Value during flooding
-------------------------------  -------------------------------  --------------------------------  --------------------------------------------------
        `r gh_min`                          `r gh_max`                   `r phi_mean`                               `r gh_phi_mean`                     
        









